---
title: "Graphics with ggplot2"
---

This .Rmd file can be composed into a Notebook result (.html or .pdf) with RStudio. Any of the R code chunks can also be submitted or be part of a batch Rscript submission.

Before we begin, note that an amazing collection of ggplot2 and other R graphics examples that go way beyond this short tutorial and include animations as well as interactive graphics is available at [The R Graph Gallery](https://www.r-graph-gallery.com/).

However, it is useful to start with some basics so The R Graph Gallery makes more sense.
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

First, we load the library.
```{r}
library(ggplot2)
```

Next, consider the `mpg` dataset that is included with the ggplot2 library.
``` {r}
tail(mpg)
```

ggplot is based on the Grammar of Graphics decomposition of graphing concepts, which allows very high-level control over graphics along with the ability to tune fine details. It is considered the gold standard for publication quality graphics. It is currently used by publications like The New York Times and The Economist due to its ability to quickly prototype custom graphics.

Grphics are decomposed into layers. Let's begin with an empty xy-plot. The aesthetics `aes()` function assigns `x = cty` and `y = hwy`.
``` {r}
g = ggplot(mpg, aes(cty, hwy))
g
```

Now we add `geom_point()` to tell ggplot how to represent the xy data. The `print()` can be ommitted as the `print()` method is the default for an object that is not assigned.
``` {r}
g = ggplot(mpg, aes(cty, hwy)) +
    geom_point()
g
```

Add another layer with a LOESS (locally estimated scatterplot smoothing) regression line and a confidence region (95% by default) for the line.  Also note that one can just add to a previously generated plot specification.
``` {r}
g + geom_smooth()
```

Customize axes, theme, etc. Notably, we choose generalized additive models from the package `mgcv` to fit thin plate splines with crossvalidation to the data instead of the default LOESS above. Again, a 95% confidence region is shown for the local mean.
``` {r}
g = ggplot(mpg, aes(cty, hwy)) +
    geom_point(shape = 21, size = 2) +
    geom_smooth(method = mgcv::gam, formula = y ~ s(x, bs = "tp")) +
    labs(x = "City", y = "Highway", title = "Miles Per Gallon") +
    theme_bw()
g
```

Now change the aesthetics to `x = hwy`, dropping the `y`, and its representation to a histogram.
``` {r}
g = ggplot(mpg, aes(hwy)) +
    geom_histogram()
g
```

Customize the histogram. Note that aesthetics are attributes that depend on data (such as the fill color below) whereas data-independent attributes (such as number of bins and outline color below) are specified in the `geom`s. 
``` {r}
g = ggplot(mpg, aes(x = hwy, fill = factor(cyl))) +
    geom_histogram(bins=20, size=0.25, color = "black")
g
```

Plot a histogram for each value of `cyl`.
``` {r}
g = ggplot(mpg, aes(hwy)) +
    geom_histogram(bins=20, fill="blue", color="black", size=0.25) +
    facet_wrap(~ cyl)
g
```

Plot boxplots for each value of `class`.
``` {r}
g = ggplot(mpg, aes(class, hwy)) +
    geom_boxplot()
g
```

Plot two boxplots together horizontally.
``` {r}
library(gridExtra)

g = ggplot(mpg, aes(class, hwy)) +
    geom_boxplot()
h = ggplot(mpg, aes(class, cty)) +
    geom_boxplot()
grid.arrange(g, h)
```

Suppose we have a time series, $y_t$, that is an autoregressive Gaussian process of order 2, denoted by AR(2):
\begin{align*}
y_t = \phi_1 y_{t-1} + \phi_2 y_{t-2} + \epsilon_t, \quad
\epsilon_t \stackrel{\text{iid}}{\sim} \text{N}(0, \sigma^2),
\quad t = 1, \ldots, n.
\end{align*}

Generate the AR(2) series and plot it.
``` {r}
y = arima.sim(n = 200, list(ar = c(0.5, -0.2), sd = sqrt(0.25)))
dat = data.frame(t = 1:200, y = as.numeric(y))
g = ggplot(dat, aes(t, y)) + geom_line()
g
```

What about plotting multiple series on one plot? First, simulate the series.
``` {r}
n = 12*60
y1 = 0 + arima.sim(n = n, list(ar = c(0.5, -0.2), sd = sqrt(0.25)))
y2 = 3 + arima.sim(n = n, list(ar = c(0.1, -0.2), sd = sqrt(0.25)))
y3 = -3 + arima.sim(n = n, list(ar = c(0.7, -0.2), sd = sqrt(0.5)))
```

Make the series the columns of a `data.frame` along with a time variable.
``` {r}
dat = data.frame(time = 1:n, y1 = as.numeric(y1), y2 = as.numeric(y2), y3 = as.numeric(y3))
head(dat, 3)
```

Gather the `data.frame` variables (except `time`) by stacking the series vertically. The `tidyr` package is part of the `tidyverse` group of packages, which include `ggplot2` as well. Note that we are not loading the `tidyr` library into our namespace and only invoking its `gather` function.
``` {r}
newdat = tidyr::gather(dat, key = "key", value = "value", -time)
head(newdat)
tail(newdat)
```

A `gather`ed data frame allows ploting the series together and using the `key` variable for data-dependent attributes.
``` {r}
g = ggplot(newdat, aes(x = time, y = value,
            group = key,
            color = key,
            linetype = key)) +
    geom_line() +
    theme(legend.position = "bottom") +
    ggtitle("Multiple Series")
g
```

If time is in minutes, perhaps the plot should be on a 12 hour polar coordinate clock?
``` {r}
g + coord_polar() + scale_x_time()
```

Draw from bivariate normal.
``` {r}
library(mvtnorm)
Sigma = matrix(c(1, 1/2, 1/2, 1), 2, 2)
x = rmvnorm(n = 10000, mean = c(0,0), sigma = Sigma)
dat = data.frame(x)
colnames(dat) = c("x", "y")
```

Plot the points and superimpose contours.
``` {r}
g = ggplot(dat, aes(x=x, y=y)) +
    geom_point(alpha = 0.2) +
    geom_density2d()
g
```

### For an amazing collection of ggplot2 and other R graphics examples that go way beyond this short tutorial and include animations as well as interactive graphics see [The R Graph Gallery](https://www.r-graph-gallery.com/).